---
title: "UMAP"
author: "Lorenzo"
date: "5/5/2019"
output: 
   github_document:
      toc: true
      toc_depth: 4
      fig_width: 5
      fig_height: 3
      
---

## Setup

```{r setup-knitr, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
```

```{r setup}
suppressPackageStartupMessages(library(tidyverse))

# Setup Python to use umap conda environment
library(reticulate)
reticulate::use_condaenv('umap', required = TRUE)
```

## R

```{r}
# Load UMAP
library(umap)

# Prepare data
iris_mtx <- iris[, c(1:4)]
```


```{r umap-plot}
# Utility to perform UMAP and plot

draw_umap <- function(data_mtx, labels, method = '', n_neighbors = 15, min_dist = 0.1) {
      
   umap_config <- umap.defaults
   umap_config$n_neighbors <- n_neighbors
   umap_config$min_dist <- min_dist
   umap_config$random_state <- 42         # Fix the seed
      
   u <- umap(data_mtx, 
             method = method, 
             config = umap_config)
   
   # Create data.frame with UMAP coordinates
   df_umap <- as.data.frame(u$layout)
   colnames(df_umap) <- c('x_umap', 'y_umap')
   df_umap <- cbind(df_umap, labels = factor(labels))
   
   p <- ggplot(df_umap, aes(x_umap, y_umap)) +
      geom_point(aes(col = labels)) +
      labs(title = glue::glue('Method: {method}, n_neighbors: {n_neighbors}, min_dist: {min_dist}'))
   
   list(p = p, df = df_umap, u = u)
}
```


### Sweep on `min_dist`

#### Naive

Calling `umap::umap` directly:

```{r naive-d-sweep}
u <- umap(iris_mtx, method="naive", min_dist = 0.001)
head(u$layout)

u <- umap(iris_mtx, method="naive", min_dist = 0.1)
head(u$layout)

u <- umap(iris_mtx, method="naive", min_dist = 0.99)
head(u$layout)

```

```{r naive-plot, echo=FALSE}
draw_umap(iris_mtx, iris$Species, method = 'naive', min_dist = 0.001)$p
draw_umap(iris_mtx, iris$Species, method = 'naive', min_dist = 0.1)$p
draw_umap(iris_mtx, iris$Species, method = 'naive', min_dist = 0.99)$p
```


#### Using Python backend

Calling `umap::umap` directly. Embeddings do not change:

```{r learn-d-sweep, collapse=TRUE}
u <- umap(iris_mtx, method="umap-learn", min_dist = 0.001)
head(u$layout)

u <- umap(iris_mtx, method="umap-learn", min_dist = 0.1)
head(u$layout)

u <- umap(iris_mtx, method="umap-learn", min_dist = 0.99)
head(u$layout)
```

```{r learn-d-plot, echo=FALSE}
draw_umap(iris_mtx, iris$Species, method = 'umap-learn', min_dist = 0.001)$p
draw_umap(iris_mtx, iris$Species, method = 'umap-learn', min_dist = 0.1)$p
draw_umap(iris_mtx, iris$Species, method = 'umap-learn', min_dist = 0.99)$p
```

Yet `min_dist` is correctly passed to `umap-learn`:

```{r learn-d-debug}
u$UMAP
u$config$umap_learn_args
```


### Sweep on `n_neighbors`

#### Naive

Calling `umap::umap` directly:

```{r naive-nn-sweep}
u <- umap(iris_mtx, method="naive", n_neighbors = 3)
head(u$layout)
u <- umap(iris_mtx, method="naive", n_neighbors = 5)
head(u$layout)
u <- umap(iris_mtx, method="naive", n_neighbors = 30)
head(u$layout)
```

```{r naive-nn-plot, echo=FALSE}
draw_umap(iris_mtx, iris$Species, method = 'naive', n_neighbors = 3)$p
draw_umap(iris_mtx, iris$Species, method = 'naive', n_neighbors = 5)$p
draw_umap(iris_mtx, iris$Species, method = 'naive', n_neighbors = 30)$p
```


#### Using Python backend

Calling `umap::umap` directly. Results are sensitive on `n_neighbors`:

```{r learn-nn-sweep}
u <- umap(iris_mtx, method="umap-learn", n_neighbors = 3)
head(u$layout)
u <- umap(iris_mtx, method="umap-learn", n_neighbors = 5)
head(u$layout)
u <- umap(iris_mtx, method="umap-learn", n_neighbors = 30)
head(u$layout)
```

```{r learn-nn-plot, echo=FALSE}
draw_umap(iris_mtx, iris$Species, method = 'umap-learn', n_neighbors = 3)$p
draw_umap(iris_mtx, iris$Species, method = 'umap-learn', n_neighbors = 5)$p
draw_umap(iris_mtx, iris$Species, method = 'umap-learn', n_neighbors = 30)$p
```

`UMAP` object is updated:

```{r learn-nn-debug}
u$UMAP
u$config$umap_learn_args
```



## Python

Calling from a `reticulate` session: no problems.

```{python py-init}
import numpy as np
from sklearn.datasets import load_iris
import matplotlib.pyplot as plt
import umap

# Prepare data
iris = load_iris()

def draw_umap(data, labels, n_neighbors=15, min_dist=0.1, title='', close=True):
   fit = umap.UMAP(
      n_neighbors=n_neighbors,
      min_dist=min_dist,
      n_components=2
   )
   
   u = fit.fit_transform(data);
   
   if (close):
      plt.close('all')
   plt.scatter(u[:,0], u[:,1], c=labels, s=1)
   plt.title('nn = {}, d = {}'.format(n_neighbors, min_dist))
   plt.show()
```

### Sweep on `min_dist`

```{python py-d-sweep}
draw_umap(iris.data, iris.target, min_dist=0.001)
draw_umap(iris.data, iris.target, min_dist=0.1)
draw_umap(iris.data, iris.target, min_dist=0.99)
```

### Sweep on `n_neighbors`

```{python py-nn-sweep}
draw_umap(iris.data, iris.target, n_neighbors=3)
draw_umap(iris.data, iris.target, n_neighbors=5)
draw_umap(iris.data, iris.target, n_neighbors=30)
```

### Python information

```{python python-info, python, echo=FALSE}
import IPython

print(IPython.sys_info())
```

Packages in current environment:

```{python pip-info, collapse=FALSE}
import subprocess

result = subprocess.run(['pip', 'list'], 
   stdout=subprocess.PIPE, 
   stderr=subprocess.PIPE, 
   stdin=subprocess.PIPE, 
   shell=True)
   
print(result.stdout.decode())
```


### R information

```{r r-info}
sessionInfo()
```

